---
title: "Compare algorithms to predict Fu_hep"
author: "Susana Proenca"
format: html
editor: visual
---

## Theory

Fraction unbound in a hepatocyte incubation (Fu_hep) is estimated using the same physicochemical partitioning principles as Fu_mic (see the "Evaluate Fu_mic" vignette for the full derivations); the hepatocyte-specific implementations differ mainly in which literature regressions apply and in the cell composition assumptions used to derive the in vitro compartment volumes.

#### Austin et al 2002 (hepatocyte form)

Uses the same functional form as the microsomal regression but with coefficients recalibrated for hepatocyte concentration (in million cells/mL) rather than microsomal protein concentration.

#### Kilford et al 2008

A hepatocyte-specific regression (there is no microsomal equivalent in this package), scaling the Hallifax and Houston microsomal correction to hepatocyte cell volume.

$$
fu_{inc} = \frac{1}{1+ 125 \cdot VR \cdot 10^{0.072\cdot \log \left( \frac{P}{D} \right)^2 +0.067\cdot\log \left( \frac{P}{D} \right)-1.126}}
$$

where VR is the cell volume to incubation volume ratio.

#### Poulin

Same equation as for microsomes, but the neutral lipid and (for strong bases) acidic phospholipid concentrations are derived from hepatocyte, not microsomal, composition.

#### Rodgers & Rowland

Same restriction as for Fu_mic: only implemented for strong bases (ionization class "base" with pKa \> 7), since it is calibrated from the compound's blood-cell partitioning (fraction unbound in plasma and blood:plasma ratio).

#### All_literature

Average of this package's own Austin, Kilford and Poulin predictions.

```{r}
#| label: load functions and data
#| include: false

library(dplyr)
library(ggpubr)
library(ggplot2)

library(ESQivive)

data_path <- system.file("extdata", "test_fu_hepatocytes.csv", package = "ESQivive")
testFuHepData <- read.csv(data_path)

# Class is coded as single letters (A/B/N) in this dataset rather than the
# full words ESQivive expects.
class_map <- c(A = "acid", B = "base", N = "neutral")
testFuHepData$Class <- class_map[testFuHepData$Class]
```

We used the Obs_Fuhep from this dataset to evaluate the predictions. Unlike the microsomal dataset, this one includes compound-specific fraction unbound in plasma (`fuP`) and blood:plasma ratio (`RBP`) for a subset of the base compounds - these are used for Poulin's strong-base branch and for Rodgers & Rowland wherever available, instead of a generic literature default.

#### Calculate Fu_hep for all

```{r}
#| echo: false

QSARs <- c(
  "Austin",
  "Kilford",
  "Poulin",
  "All Poulin and Theil",
  "All Berezhkovskiy",
  "All PK-Sim Standard",
  "Rodgers & Rowland + fu",
  "All Schmitt",
  "All_literature"
)
QSAR_colnames <- c(
  "ESQivive_Austin",
  "ESQivive_Kilford",
  "ESQivive_Poulin",
  "All_Poulin_and_Theil",
  "All_Berezhkovskiy",
  "All_PK_Sim_Standard",
  "Rodgers_Rowland_fu",
  "All_Schmitt",
  "All_literature"
)

# Rodgers & Rowland only supports strong bases (see Theory above) and, along
# with Poulin's strong-base branch and All_literature, needs fraction unbound
# in plasma / blood:plasma ratio - taken from this dataset's fuP/RBP columns
# where available (NA otherwise, which safe_fu() turns into a clean NA result
# rather than an error).
safe_fu <- function(...) tryCatch(calculate_fu_in_vitro(...), error = function(e) NA_real_)

for (j in seq_along(QSARs)) {
  testFuHepData[[QSAR_colnames[j]]] <- sapply(seq_len(nrow(testFuHepData)), function(i) {
    # pKa is not numeric for neutral compounds (no ionizable group); this is
    # fine since none of the QSARs above use pka when the compound is neutral.
    suppressWarnings(safe_fu(
      partition_qspr = QSARs[j],
      log_lipophilicity = as.double(testFuHepData[i, "LogP37C"]),
      ionization = c(testFuHepData[i, "Class"], 0),
      type_system = "hepatocytes",
      FBS_fraction = 0,
      microplate_type = 96,
      volume_medium = 0.22,
      pka = c(as.double(testFuHepData[i, "pKa"]), 0),
      concentration_cells = as.double(testFuHepData[i, "Cell_concentration.10_6cells.mL."]),
      blood_plasma_ratio = as.double(testFuHepData[i, "RBP"]),
      fraction_unbound = as.double(testFuHepData[i, "fuP"])
    ))
  })
}

cat(
  "Rodgers & Rowland computed for", sum(!is.na(testFuHepData$Rodgers_Rowland_fu)),
  "of", nrow(testFuHepData), "compounds (strong bases with known fuP/RBP only)\n"
)

head(testFuHepData)
```

This dataset also gives two literature-reported prediction columns (`Fuhep_Poulin`, `Fu_Austin`), letting us sanity-check this package's own `ESQivive_Poulin`/`ESQivive_Austin` against the originally reported values.

```{r}
plot(testFuHepData$Fuhep_Poulin, testFuHepData$ESQivive_Poulin,
  xlab = "Fuhep_Poulin (paper)", ylab = "ESQivive_Poulin (this package)"
)
abline(0, 1, col = "red")
```

```{r}
plot(testFuHepData$Fu_Austin, testFuHepData$ESQivive_Austin,
  xlab = "Fu_Austin (paper)", ylab = "ESQivive_Austin (this package)"
)
abline(0, 1, col = "red")
```

Poulin agrees closely with the paper's own values (points fall near the identity line). Austin does not - this package's `calculate_fu_hep_austin()` systematically predicts lower Fu than the paper's `Fu_Austin` column, which is worth investigating further (e.g. whether the coefficients or cell-concentration assumptions used here match the original hepatocyte-specific regression) before relying on it.

Verapamil appears twice in this dataset (rat and human), giving a small comparison between the spread across our prediction methods and the spread between the two species' observed values.

```{r}
verapamil_rows <- testFuHepData[testFuHepData$Compound == "Verapamil", ]
human_row <- verapamil_rows[verapamil_rows$Species == "Human", ]
predicted_cols <- c("Fuhep_Poulin", "Fu_Austin", QSAR_colnames)

verapamil_df <- data.frame(
  fu = c(as.double(human_row[1, predicted_cols]), verapamil_rows$Obs_Fuhep),
  type = c(
    rep("predicted (all methods, human)", length(predicted_cols)),
    rep("observed (rat vs. human)", nrow(verapamil_rows))
  )
)

ggplot(verapamil_df, aes(x = type, y = fu, color = type)) +
  geom_jitter(width = 0.05, height = 0, size = 3) +
  labs(x = NULL, y = "Fu_hep", title = "Verapamil") +
  theme_bw() +
  theme(legend.position = "none")
```

### Plots

```{r}
#| fig.width: 12
#| fig.height: 10

# First 2 columns are the original papers' own reported predictions (from the
# dataset itself); the rest are computed by this package above.
plot_cols <- c("Fuhep_Poulin", "Fu_Austin", QSAR_colnames)

make_fu_plot <- function(data, ycol) {
  ggplot(data, aes(x = Obs_Fuhep, y = .data[[ycol]], col = Class)) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_point() +
    theme_bw() +
    labs(title = ycol, y = "Predicted Fu_hep") +
    stat_regline_equation(aes(label = after_stat(rr.label)))
}

fu_plots <- lapply(plot_cols, make_fu_plot, data = testFuHepData)
ggarrange(plotlist = fu_plots, common.legend = TRUE)
```

### Error Table

For each method and each chemical class (acid/base/neutral), we calculate:

-   `percent_within_2fold` / `percent_within_5fold`: the percentage of predictions within 2-fold / 5-fold of the observed value
-   `afe_method` (Average Fold Error) and `aafe_method` (Average Absolute Fold Error): overall magnitude of error, direction-agnostic
-   `bias_fold`: the signed geometric-mean fold error - values \>1 indicate the method tends to over-predict on average, \<1 under-predict
-   `rmse`: root mean squared error, on the same 0-1 scale as Fu
-   `r2`: Pearson r², the strength of the *linear* relationship between predicted and observed
-   `spearman_rho`: Spearman rank correlation, robust to nonlinearity and outliers
-   `ccc`: Lin's Concordance Correlation Coefficient - penalizes systematic deviation from the line of identity (y = x), not just correlation

Rodgers & Rowland only has predictions for strong bases with known fuP/RBP, so all metrics for that method are computed only over its available (non-`NA`) subset - `n` reports how many compounds contributed to each row.

```{r}
# All metrics below are computed only on rows where both observed and
# predicted are finite, since Rodgers & Rowland only has results for a subset
# of strong bases and would otherwise silently drop out of every summary.
complete_pairs <- function(observed, predicted) {
  ok <- is.finite(observed) & is.finite(predicted)
  list(observed = observed[ok], predicted = predicted[ok])
}

percent_within_fold <- function(observed, predicted, fold) {
  cp <- complete_pairs(observed, predicted)
  mean(cp$predicted >= cp$observed / fold & cp$predicted <= cp$observed * fold) * 100
}

afe <- function(observed, predicted) {
  cp <- complete_pairs(observed, predicted)
  mean(abs(cp$predicted / cp$observed))
}

aafe <- function(observed, predicted) {
  cp <- complete_pairs(observed, predicted)
  mean(abs(log10(cp$predicted / cp$observed)))
}

bias_fold <- function(observed, predicted) {
  cp <- complete_pairs(observed, predicted)
  10^(mean(log10(cp$predicted / cp$observed)))
}

rmse <- function(observed, predicted) {
  cp <- complete_pairs(observed, predicted)
  sqrt(mean((cp$predicted - cp$observed)^2))
}

pearson_r2 <- function(observed, predicted) {
  cp <- complete_pairs(observed, predicted)
  if (length(cp$observed) < 3) {
    return(NA_real_)
  }
  cor(cp$observed, cp$predicted, method = "pearson")^2
}

spearman_rho <- function(observed, predicted) {
  cp <- complete_pairs(observed, predicted)
  if (length(cp$observed) < 3) {
    return(NA_real_)
  }
  cor(cp$observed, cp$predicted, method = "spearman")
}

lin_ccc <- function(observed, predicted) {
  cp <- complete_pairs(observed, predicted)
  if (length(cp$observed) < 2) {
    return(NA_real_)
  }
  mean_o <- mean(cp$observed)
  mean_p <- mean(cp$predicted)
  var_o <- mean((cp$observed - mean_o)^2)
  var_p <- mean((cp$predicted - mean_p)^2)
  covar <- mean((cp$observed - mean_o) * (cp$predicted - mean_p))
  (2 * covar) / (var_o + var_p + (mean_o - mean_p)^2)
}

error_table <- list()

for (predi in plot_cols) {
  error_table[[predi]] <- testFuHepData %>%
    group_by(Class) %>%
    summarize(
      n = sum(is.finite(.data[[predi]])),
      percent_within_2fold = percent_within_fold(Obs_Fuhep, .data[[predi]], 2),
      percent_within_5fold = percent_within_fold(Obs_Fuhep, .data[[predi]], 5),
      afe_method = afe(Obs_Fuhep, .data[[predi]]),
      aafe_method = aafe(Obs_Fuhep, .data[[predi]]),
      bias_fold = bias_fold(Obs_Fuhep, .data[[predi]]),
      rmse = rmse(Obs_Fuhep, .data[[predi]]),
      r2 = pearson_r2(Obs_Fuhep, .data[[predi]]),
      spearman_rho = spearman_rho(Obs_Fuhep, .data[[predi]]),
      ccc = lin_ccc(Obs_Fuhep, .data[[predi]]),
      .groups = "drop"
    )
}

# error_table[["All_literature"]]
```
